Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-08-19 docs/submission-mee-2020-08-19.tex
second draft final :tada:

\textbf{Word count}: 3775

@@

The prevalence of taxonomic idiosyncrasies across databases represent a key challenge to automatically connecting data from disparate biological databases in a phylogenetic context. To standardize taxonomic names, a unified system is needed. The main aim of the Open Tree of Life (OpenTree from now on) project
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's ``synthetic'' tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies taxonomy from various databases (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}) such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}; Wheeler \emph{et al.} \protect\hyperlink{ref-wheeler2000database}{2000}), the Global Biodiversity Information Facility (GBIF; Secretariat \protect\hyperlink{ref-secretariat2017gbif}{2017}), and the World Register of Marine Species {[}WoRMS; www.marinespecies.org/{]}, providing a key resource that can be used to connect data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses.
Species tree reconstructions from multiple single locus data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships (Song \emph{et al.} \protect\hyperlink{ref-song2012resolving}{2012}).
Genomics has, and will continue to, revolutionize phylogenetic inference.
Yet, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing, single nucleotide polymorphisms, microsatellites, and ultra-conserved elements, which has lead to largely non-overlapping genomic data sets across taxa, difficulting wide scale phylogenetic reconstructions.
While phylogenomics ameliorates the problem of non-overlapping genomic data sets by focusing on targeted capture of informative characters from independent and single-copy genetic markers (Jones \& Good \protect\hyperlink{ref-jones2016targeted}{2016}; Andermann \emph{et al.} \protect\hyperlink{ref-andermann2020guide}{2020}), decades of single locus sequencing have already generated homologous DNA data sets that can be used for phylogenetic reconstruction at many scales.

Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences. These sequences were gauged to have the potential to resolve phylogenetic relationships of most (98.05\%) of the almost 241, 000 distinct taxa in the NCBI taxonomy at the time (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}). Assembling a DNA alignment from such a massive database can be done ``by hand'', but it requires huge amounts of time and it is mostly a non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to study evolutionary relationships among a variety of organisms (e.g., Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}; Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}; Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017}).
However, even during phylogenomic reconstructions, thoughtfully curated markers and alignments can improve phylogenetic reconstructions (Fragoso-Martínez \emph{et al.} \protect\hyperlink{ref-fragoso2017pilot}{2017}).

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as ``jump-start'' alignments (Morrison \protect\hyperlink{ref-morrison2006multiple}{2006}). The TreeBASE database (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa \href{https://www.treebase.org/treebase-web/home.html\#:~:text=TreeBASE\%20is\%20produced\%20and\%20governed,mapped\%20to\%20104\%2C593\%20distinct\%20taxa.}{(see TreeBASE's website about)}, representing a source of valuable expert knowledge. Linking published alignments with public molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life. The phylogenies associated to TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}), and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.

Physcraper relies on programmatic access protocols available to automatically link molecular data from GenBank to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to continually update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for comparison of existing phylogenetic hypotheses with newly generated ones.
Physcraper is coded as a Python pipeline that focuses on data interoperability, by integrating taxonomic name matching across biological databases. This integration also allows users to rapidly place new data from a diverse range of biological databases in an evolutionary context, making a variety of downstream analyses straightforward.

@@

\begin{itemize}
\item
  In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree \href{https://github.com/opentreeoflife/phylesystem}{Phylesystem}, or submitted via OpenTree's curator \href{https://tree.opentreeoflife.org/curator}{application} (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}).
  Currently, only trees connected to a published study can be stored in the Phylesystem.
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on TreeBASE.
  If the user is not ready to make the input tree public, tree tip labels must be first standardized to the unified OpenTree taxonomy using OpenTree's bulk Taxonomic Name Resolution Service \href{https://tree.opentreeoflife.org/curator/tnrs/}{TNRS} tool. \emph{Should we say that it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?}
  This step is referred to as taxonomic name mapping and Phylesystem stored trees are processed in this way upon submission.
  Physcraper saves a summary ``csv'' file with results from the taxon name standardization for advantage of the user.
  If taxon names can't be mapped, their taxonomic information can't be used in comparison analysis. They will still be used in the sequence search and phylogenetic reconstruction steps.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the standardized taxonomic names and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline.
\item
  The input alignment should be a single locus alignment that was used to generate the tree. Alignments are often stored in a public repository such as TreeBase (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}; Vos \emph{et al.} \protect\hyperlink{ref-vos2012nexml}{2012}),
  DRYAD (\href{http://datadryad.org/}{www.datadryad.org}), or a data repository associated with the journal where the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper downloads it directly,
  either from the TreeBASE website (\href{https://treebase.org/}{www.treebase.org})
  or through the TreeBASE GitHub repository (SuperTreeBASE; \href{https://github.com/TreeBASE/supertreebase}{github.com/TreeBASE/supertreebase}).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
\item
  Single locus alignments sometimes have fewer taxa than the tree inferred from the full concatenated data, simply because a single molecular marker usually does not cover all the taxa sampled for the full phylogenetic analysis. Physcraper prunes the input tree to taxa found in the alignment, and verifies that all taxon names on the tips of the tree are in the DNA character matrix and vice versa. Technically, just one taxon name (and its corresponding sequence in the alignment) is needed to continue the algorithm. See next section.
\item
  The standardized and pruned tree and alignment (checked tree and alignment from now on) are output as ``newick'' and ``fasta'' respectively in the ``inputs'' folder to be used in the following steps.
\end{itemize}

@@

\begin{itemize}
\item
  Physcraper uses the GenBank DNA database as source to search for new sequences. The DNA sequence search can be performed on the GenBank remote database or in a GenBank local database set up by the user, which can speed up the search process. Detailed instructions to setup a local database are provided on Physcraper's software documentation.
\item
  The next step is to identify a ``search taxon'' to constrain the sequence search on the GenBank database within that taxonomic group.
  The search taxon can be chosen by the user from the NCBI taxonomy.
  If none is provided, then the search taxon is automatically set using the taxa in the input tree labeled as the ``ingroup'' (Fig. \ref{fig:framework}).
  The search taxon is The Most Recent Common Ancestor (MRCA) of the ingroup taxa in the OpenTree synthetic tree, that is also a named clade in the NCBI taxonomy.
  This is known in the OpenTree as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred as the Least Inclusive Common Ancestral taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.
  To identify the MRCAT of a group of taxon names, we use the OpenTree \href{https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3\#mrca}{taxonomic tool v3} (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}).

  Users can provide a search taxon that is either a more or a less inclusive
  clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.
\item
  The Basic Local Alignment Search Tool, BLAST (Altschul \emph{et al.} \protect\hyperlink{ref-altschul1990basic}{1990}, \protect\hyperlink{ref-altschul1997gapped}{1997}) is used to identify
  similarity between DNA sequences within the search taxon in a nucleotide
  database, and the sequences on the checked alignment.
  The \texttt{blastn} function from the BLAST command line tools (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) is used for local database sequence searches.
  For remote database searches, we modified the BioPython (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}) BLAST function from the \href{https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html}{NCBIWWW module} to accept an alternative BLAST address (URL). This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST avoiding NCBI's required waiting times, which slow down the searches markedly.
\item
  A constrained BLAST search is performed, in which each sequence
  in the alignment is BLASTed once against all database DNA sequences belonging to the search
  taxon. All results from each BLAST run are stored, and sequences with match scores better than the e-value cutoff (default to 0.00001) are saved
  along with their corresponding metadata, i.e., their GenBank accession number.
  The full sequence for each match is downloaded from NCBI into a dedicated library within the ``physcraper'' folder, allowing for secondary analyses to run significantly faster.
\item
  BLAST result sequences will be discarded if they fall outside the user set min and max length cutoffs, set as proportions of the average length without gaps of sequences in the input alignment (defaults values of 80\% and 120\%, respectively).
  This filtering guarantees the exclusion of whole genome sequences, which create problems in multiple sequence alignment.
  The GenBank accession numbers of sequences removed due to not meeting e-value or length cutoffs are stored in output files.
\item
  All sequences accepted up to this point are assigned an internal identifier, and are further filtered.
\item
  New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the OTT taxonomy or the NCBI taxonomy, or they are longer than the sequence in the input alignment.
\item
  Among the filtered sequences, there are often several representatives per taxon.
  Although it can be useful to keep some of them, for example, to investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon in downstream analyses,
  5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.
\item
  All BLAST and filtering parameters can be customized by the user.
\item
  Reverse, complement, and reverse-complement BLAST result sequences are identified and translated using BioPython internal functions (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}).
\item
  Iterative cycles of sequence similarity search can be performed, by blasting the newly found sequences until no new sequences are found. By default only one BLAST search cycle is performed in which only sequences in the input alignment are blasted.
\item
  New sequences passing all filtering steps are added to the ``csv'' taxon summary file.
\item
  A ``fasta'' file containing all new filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  By default, Physcraper uses the software MUSCLE (Edgar \protect\hyperlink{ref-edgar2004muscle}{2004}) to perform DNA sequence alignments. Instructions on how to install all software dependencies used by Physcraper are provided in the documentation.
\item
  The process to align new sequences consists of two steps. First, all new sequences are aligned using the default MUSCLE options.
\item
  Second, a MUSCLE profile alignment is performed, in which the original alignment
  is used as a template to align the new sequences. This ensures that the final alignment
  follows the homology criteria established by the original alignment.
\item
  The final alignment is not further processed by Physcraper. It is recommended that the alignment is checked by the user, by eye followed by manual refinement, or using a tool for automatic alignment processing (e.g., GBlocks; Castresana \protect\hyperlink{ref-castresana2000selection}{2000}, \protect\hyperlink{ref-castresana2002gblocks}{2002}).
\item
  While curating the alignment is a critical step, it is not a reproducible one. The main reason for its lack of reproducibility might be that it is hard to track changes made on the alignment. A form of version control, to register the differences between the alignment that was produced by the software and the manually curated alignment will be ideal.
\item
  Users may also use Physcraper to only gather new GenBank sequences, to then apply their own preferred alignment and phylogenetic inference methods.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  A Maximum Likelihood (ML) gene tree is reconstructed for each alignment provided, using the software RAxML (Stamatakis \protect\hyperlink{ref-stamatakis2014raxml}{2014}) with default settings, such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with the default algorithm. Currently only the number of bootsrap replicates can be specified by the user.
\item
  By default, the original tree is used as a starting tree for the ML searches. Alternatively, users can set the original tree as a full topological constraint, or ignore it completely for the searches.
\item
  Bootstrap results are summarized with the SumTrees module of DendroPy (current version 4.4.0; Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}).
\item
  Physcraper's final result is an updated phylogenetic hypothesis for the locus provided in the input alignment.
\item
  Tips on all trees generated by Physcraper are defined by a taxon ``name space''. The taxon metadata captures the NCBI accession information, as well as the taxon identifiers, allowing the user to perform comparisons and conflict analyses.
\item
  Two ways to compare the updated tree with the original tree are implemented in Physcraper. First, Robinson Foulds weighted and unweighted metrics are estimated using Dendropy functions (Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}).
\item
  Second, a conflict analysis is performed. This is a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface (Redelings \& Holder \protect\hyperlink{ref-redelings2017supertree}{2017}).
\item
  For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined.
\item
  A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroup directly on the updated tree, so trees are accurately rooted.
\end{itemize}

@@

Data repositories designed to preserve and democratize access to biological data
constitute an essential resource for evolutionary estimation. While data keep accumulating,
our ability to reanalyze and incorporate new data to available knowledge is being challenged.

Various phylogenetic pipelines have been designed to make sense of the vast amount of public molecular data available in different ways.
Pipelines like Phylota (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}), PyPHLAWD (Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}), and SUPERSMART (Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017})
focus mainly on generating full trees \emph{de novo} for regions of the Tree of
Life that have no phylogenetic assessment yet in published studies.
While Physcraper does not generate phylogenies \emph{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of pre-existing
phylogenetic knowledge available in tree repositories to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline (Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}) also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to conduct a comparison analysis to Physcraper.

>>>>

\textbf{Word count}: 3911

@@

The prevalence of taxonomic idiosyncrasies across databases represent a key challenge to automatically connecting data from disparate biological databases in a phylogenetic context. To standardize taxonomic names, a unified system is needed. The main aim of the Open Tree of Life project (OpenTree)
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's ``synthetic'' tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies taxonomy from various databases (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}) such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}; Wheeler \emph{et al.} \protect\hyperlink{ref-wheeler2000database}{2000}), the Global Biodiversity Information Facility (GBIF; Secretariat \protect\hyperlink{ref-secretariat2017gbif}{2017}), and the World Register of Marine Species {[}WoRMS; www.marinespecies.org/{]}, providing a key resource that can be used to connect data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses.
Species tree reconstructions from multiple single locus data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships (Song \emph{et al.} \protect\hyperlink{ref-song2012resolving}{2012}).
Genomics has, and will continue to, revolutionize phylogenetic inference.
Yet, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing, single nucleotide polymorphisms, microsatellites, and ultra-conserved elements, which has lead to largely non-overlapping genomic data sets across taxa, creating difficulty in wide scale phylogenetic reconstructions.
While phylogenomics ameliorates the problem of non-overlapping genomic data sets by focusing on targeted capture of informative characters from independent and single-copy genetic markers (Jones \& Good \protect\hyperlink{ref-jones2016targeted}{2016}; Andermann \emph{et al.} \protect\hyperlink{ref-andermann2020guide}{2020}), decades of single locus sequencing have already generated homologous DNA data sets that can be used for phylogenetic reconstruction at many scales.

Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million DNA sequences. These sequences were gauged to have the potential to resolve phylogenetic relationships of most (98.05\%) of the almost 241, 000 distinct taxa in the NCBI taxonomy at the time (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}). Assembling a DNA alignment from such a massive database can be done ``by hand'', but it requires huge amounts of time and it is mostly a non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to study evolutionary relationships among a variety of organisms (e.g., Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}; Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}; Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017}).
However, even in phylogenomic reconstructions, thoughtfully curated markers and alignments can improve phylogenetic reconstructions (Fragoso-Martínez \emph{et al.} \protect\hyperlink{ref-fragoso2017pilot}{2017}).

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as ``jump-start'' alignments (Morrison \protect\hyperlink{ref-morrison2006multiple}{2006}). The TreeBASE database (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa \href{https://www.treebase.org/treebase-web/home.html\#:~:text=TreeBASE\%20is\%20produced\%20and\%20governed,mapped\%20to\%20104\%2C593\%20distinct\%20taxa.}{(see TreeBASE's website about)}, representing a source of valuable expert knowledge. Linking published alignments with public molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life. The phylogenies associated with TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}), and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.

Physcraper relies on programmatic access protocols (API's) available to automatically link molecular data from GenBank to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to continually update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for comparison of existing phylogenetic hypotheses with newly generated ones.
Physcraper is coded as a Python pipeline that focuses on data interoperability, by integrating taxonomic name matching across biological databases. This integration also allows users to rapidly place new data from a diverse range of biological databases in an evolutionary context, making a variety of downstream analyses straightforward.

@@

In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
OpenTree \href{https://github.com/opentreeoflife/phylesystem}{Phylesystem}, or submitted via OpenTree's curator \href{https://tree.opentreeoflife.org/curator}{application} (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}).
Currently, only trees connected to a published study can be stored in the Phylesystem.
Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on TreeBASE.
If the user is not ready to make the input tree public, tree tip labels can be standardized to the unified OpenTree taxonomy using OpenTree's bulk Taxonomic Name Resolution Service \href{https://tree.opentreeoflife.org/curator/tnrs/}{TNRS} tool.
This step is referred to as taxonomic name mapping and Phylesystem stored trees are processed in this way upon submission.
Physcraper saves a summary ``csv'' file with results from the taxon name standardization for advantage of the user, including the mappings to unique identifiers in the OpenTree and NCBI taxonomies.
If taxon names can't be mapped, their taxonomic information is not used in comparison analysis. These taxa will still be used in the sequence search and phylogenetic reconstruction steps.
Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
that are essential for automatizing the phylogeny updating process. The most relevant of these is the standardized taxonomic names and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline.

The input alignment should be a single locus alignment that was used in part or in whole, to generate the tree. Alignments are often stored in a public repository such as TreeBase (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}; Vos \emph{et al.} \protect\hyperlink{ref-vos2012nexml}{2012}),
DRYAD (\href{http://datadryad.org/}{www.datadryad.org}), or a data repository associated with the journal where the tree was originally published.
If the alignment is stored in TreeBase, Physcraper can download it directly,
either from the TreeBASE website (\href{https://treebase.org/}{www.treebase.org})
or through the TreeBASE GitHub repository (SuperTreeBASE; \href{https://github.com/TreeBASE/supertreebase}{github.com/TreeBASE/supertreebase}).
If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.

Single locus alignments sometimes have fewer taxa than the tree inferred from the full concatenated data, simply because a single molecular marker usually does not cover all the taxa sampled for the full phylogenetic analysis. Physcraper prunes the input tree to taxa found in the alignment, and verifies that all taxon names on the tips of the tree are in the DNA character matrix and vice versa. Technically, just one taxon name (and its corresponding sequence in the alignment) is needed to continue the algorithm.
The standardized and pruned tree and alignment (checked tree and alignment from now on) are output as ``newick'' and ``fasta'' respectively in the ``inputs'' folder to be used in the following steps.

@@

Physcraper uses the GenBank DNA database as source to search for new sequences. The DNA sequence search can be performed on the GenBank remote database or in a GenBank local database set up by the user, which can speed up the search process. Detailed instructions to setup a local database are provided on Physcraper's software documentation.

The next step is to identify a ``search taxon'' to constrain the sequence search on the GenBank database within that taxonomic group.
The search taxon can be chosen by the user from the NCBI taxonomy.
If none is provided, then the search taxon is automatically set using the taxa in the input tree labeled as the ``ingroup'' (Fig. \ref{fig:framework}).
The search taxon is The Most Recent Common Ancestor (MRCA) of the ingroup taxa in the OpenTree synthetic tree, that is also a named clade in the NCBI taxonomy.
This is known in the OpenTree as the Most Recent Common
Ancestral Taxon (MRCAT; also referred as the Least Inclusive Common Ancestral taxon - LICA).
The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.
To identify the MRCAT of a group of taxon names, we use the OpenTree \href{https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3\#mrca}{taxonomic tool v3} (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}).

Users can provide a search taxon that is either a more or a less inclusive
clade relative to the ingroup of the original phylogeny. If the search taxon is more inclusive, the sequence search will be performed outside the MRCAT of the matched taxa, e.g., including all taxa within
the family or the order that the ingroup belongs to. If the search taxon is a less inclusive clade, the users can focus on enriching a particular clade/region within the ingroup of the phylogeny.

The Basic Local Alignment Search Tool, BLAST (Altschul \emph{et al.} \protect\hyperlink{ref-altschul1990basic}{1990}, \protect\hyperlink{ref-altschul1997gapped}{1997}) is used to identify
similarity between DNA sequences within the search taxon in a nucleotide
database, and the sequences on the checked alignment.
The \texttt{blastn} function from the BLAST command line tools (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) is used for local database sequence searches.
For remote database searches, we modified the BioPython (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}) BLAST function from the \href{https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html}{NCBIWWW module} to accept an alternative BLAST address (URL). This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST avoiding NCBI's required waiting times, which slow down the searches markedly.
A constrained BLAST search is performed, in which each sequence
in the alignment is BLASTed once against all database DNA sequences belonging to the search
taxon. All results from each BLAST run are stored, and sequences with match scores better than the e-value cutoff (default to 0.00001) are saved
along with their corresponding metadata, i.e., their GenBank accession number.
The full sequence for each match is downloaded from NCBI into a dedicated library within the ``physcraper'' folder, allowing for secondary analyses to run significantly faster.

BLAST result sequences will be discarded if they fall outside the user set min and max length cutoffs, set as proportions of the average length without gaps of sequences in the input alignment (defaults values of 80\% and 120\%, respectively).
This filtering guarantees the exclusion of whole genome sequences, which create problems in multiple sequence alignment.
The GenBank accession numbers of sequences removed due to not meeting e-value or length cutoffs are stored in output files.
All sequences accepted up to this point are assigned an internal identifier.
New sequences that are either identical or a subset of any existing sequence in the input alignment are discarded, unless they represent a different taxon in the OTT taxonomy or the NCBI taxonomy, or they are longer than the sequence in the input alignment.
Among the filtered sequences, there are often several representatives per taxon.
Although it can be useful to keep some of them, for example, to investigate monophyly
within species, there can be hundreds of exemplar sequences per taxon for some markers.
To control the number of sequences per taxon in downstream analyses,
5 sequences per taxon are chosen at random. This number is set by default but can be modified by the user.

All BLAST and filtering parameters can be customized by the user.
Reverse, complement, and reverse-complement BLAST result sequences are identified and translated using BioPython internal functions (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}).
Iterative cycles of sequence similarity search can be performed, by blasting the newly found sequences until no new sequences are found. By default only one BLAST search cycle is performed in which only sequences in the input alignment are blasted.
New sequences passing all filtering steps are added to the ``csv'' taxon summary file.
A ``fasta'' file containing all new filtered and processed sequences resulting from the BLAST search is generated for the user, and is used as an input for alignment.

@@

By default, Physcraper uses the software MUSCLE (Edgar \protect\hyperlink{ref-edgar2004muscle}{2004}) to perform DNA sequence alignments. Instructions on how to install all software dependencies used by Physcraper are provided in the documentation.
The process to align new sequences consists of two steps. First, all new sequences are aligned using the default MUSCLE options.

Second, a MUSCLE profile alignment is performed, in which the original alignment
is used as a template to align the new sequences. This ensures that the final alignment
follows the homology criteria established by the original alignment.
The final alignment is not further processed by Physcraper. It is recommended that the alignment is checked by the user, by eye followed by manual refinement, or using a tool for automatic alignment processing (e.g., GBlocks; Castresana \protect\hyperlink{ref-castresana2000selection}{2000}, \protect\hyperlink{ref-castresana2002gblocks}{2002}).
While curating the alignment is a critical step, it is not a reproducible one. The main reason for its lack of reproducibility might be that it is hard to track changes made on the alignment. A form of version control, to register the differences between the alignment that was produced by the software and the manually curated alignment would ideal.
Users may also use Physcraper to only gather new GenBank sequences, to then apply their own preferred alignment and phylogenetic inference methods.

@@

A Maximum Likelihood (ML) gene tree is reconstructed for each alignment provided, using the software RAxML (Stamatakis \protect\hyperlink{ref-stamatakis2014raxml}{2014}) with default settings, such as a GTRCAT model of molecular evolution and 100 bootstrap replicates with the default algorithm. Currently only the number of bootsrap replicates can be specified by the user.
By default, the original tree is used as a starting tree for the ML searches. Alternatively, users can set the original tree as a full topological constraint, or ignore it completely for the searches.
Bootstrap results are summarized with the SumTrees module of DendroPy (current version 4.4.0; Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}).

Physcraper's final result is an updated phylogenetic hypothesis for the locus provided in the input alignment.
Tips on all trees generated by Physcraper are defined by a taxon ``name space''. The taxon metadata captures the NCBI accession information, as well as the taxon identifiers, allowing the user to perform comparisons and conflict analyses.
Two ways to compare the updated tree with the original tree are implemented in Physcraper. First, Robinson Foulds weighted and unweighted metrics are estimated using Dendropy functions (Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}).
Second, a conflict analysis is performed. This is a node by node comparison between the the synthetic OpenTree and the original and updated tree individually. This is performed with OpenTree's conflict Application Programming Interface (Redelings \& Holder \protect\hyperlink{ref-redelings2017supertree}{2017}).
For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined. A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define outgroup directly on the updated tree, so trees are accurately rooted.

@@

Data repositories designed to preserve and democratize access to biological data
constitute an essential resource for evolutionary estimation. While data keep accumulating,
our ability to reanalyze and incorporate new data with available knowledge is being challenged.
